\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2013-12-23 fluB.tex
Adding more text here and there.
Mostly focus on deltaTMRCA and SPR distance discussion and adding more
references to B/Waikato/6/2005-like isolates (with different lineage
HA).

\subsection*{Limitations of current study}
The methods used in this paper have limited power to detect reassortments, since lineages have been fixed over time and thus all viruses eventually come to bear either Vic or Yam lineage PA, NP, NA, MP and NS segments.
We can also only detect reassortment events where a lineage has switched from Vic to Yam or \textit{vice versa}, but not when a lineage has switched from a Yam lineage to another Yam lineage, which has occurred in the past (e.g. the reassortment events giving rise to B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like viruses, which involved a Vic lineage PB1-PB2-HA complex reassorting with NP of Yamagata lineage is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, already had a Yamagata lineage-derived NP).
However, this study focuses on PB1, PB2 and HA segments, all of which have survived as two lineages up to this day and thus are more informative for inferring reassortments.

\subsection*{PB1-PB2-HA co-adaptation has evolved over time}
Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 3 genome constellations having mixed-lineage PB1-PB2-HA sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.

Our results also suggested that all segments have spent very little evolutionary time in the same virions as the 3 mixed-lineage PB1-PB2-HA constellations, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
We interpret lack of reassortment within the PB1-PB2-HA complex across the lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.

We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.

@@

\subsection*{All segments reassort equally but some are more equal than others}
The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in reassortment rate between all segments.
However, SPR distances would only approximate the true number of reassortments.
Together with an approximation of SPR distance, this might be an indication of lack of power to detect differences in reassortment rate.

@@

@@

Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.

\subsection*{The future of influenza B viruses}
We see two potential paths of evolution for influenza B viruses.
More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.

This may seem to be the case in Figure \ref{tmrcaOT} where the NP segment, exhibiting the fourth oldest TMRCA by 2011, looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes (in fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains).
The other alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes, or even extinction of one or the other lineage of the PB1-PB2-HA complex.

Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the history of influenza B viruses.
If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages get recruited into co-adapted segment complexes.

@@

@@

\section*{Conclusions}
In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
Our findings reveal that Vic and Yam PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex that co-circulate within the influenza B virus population, occasionally exchanging non-PB1-PB2-HA segments.
Though we detect strains with mixed-lineage PB1-PB2-HA complexes early after the split between Vic and Yam lineages, we find that they have performed poorly and have circulated for short amounts of evolutionary time when compared to viruses with `pure' lineage PB1-PB2-HA complexes.

We argue that this is evidence for a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
In addition, the close timing of these mixed-lineage PB1-PB2-HA reassortants close to the split of Vic and Yam lineages suggests that the relationship between each lineage's PB1-PB2-HA complexes have developed over time and in their current state reassortant viruses with mixed-lineage PB1-PB2-HA complexes are undetected by surveillance due to either poor fitness or inviability.

>>>>

\subsection*{Limitations of current study}
The methods used in this paper have limited power to detect reassortments, since lineages have been fixed over time and thus all viruses eventually come to bear either Vic or Yam lineage PA, NP, NA, MP and NS segments with the exception of PB1, PB2 and HA segments.
We can also only detect reassortment events where a lineage has switched from Vic to Yam or \textit{vice versa}, but not when a lineage has switched from a Yam lineage to another Yam lineage, which has occurred in the past (e.g. the reassortment events giving rise to B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like viruses in Figure \ref{railroadPlot}, which involved a Vic lineage PB1-PB2-HA complex reassorting with NP of Yamagata lineage is only visible in the NP tree, since all viruses in our dataset from 1995 onwards, including Vic PB1-PB2-HA complex-bearing ones, already had a Yamagata lineage-derived NP).
However, this study focuses on PB1, PB2 and HA segments, all of which have survived as two lineages up to this day and thus are more informative for inferring reassortments.

\subsection*{PB1-PB2-HA co-adaptation has evolved over time}
Our findings suggest that PB1, PB2 and HA segments of Vic and Yam lineages represent co-adapted co-reassorting segment complexes.
Previous studies have suggested similarities between the evolution of PB1, PB2 and HA segments of influenza B viruses \cite{hiromoto2000,lindstrom1999}, based on topology similarities between phylogenetic trees of the three segments, albeit the trees used contained limited numbers of sequences sampled somewhat close in time to the split of Victoria and Yamagata lineages.
Despite the apparent similarities between PB1, PB2 and HA trees, we have detected at least 4 genome constellations having mixed-lineage PB1-PB2-HA, 3 of which were sampled early in the first decade of the split between Vic and Yam lineages, when, presumably, Vic and Yam lineages of PB1, PB2 and HA segments were still similar at the amino acid level to produce viable, though presumably unfit, influenza B viruses.
The fourth genome constellation having mixed-lineage PB1-PB2-HA complex was isolated in Australia and New Zealand in 2005 over a 5 month period and have not been detected since.

Our results also suggested that all segments have spent very little evolutionary time with the 3 mixed-lineage PB1-PB2-HA constellations, compared to either entirely Yam PB1-PB2-HA or entirely Vic PB1-PB2-HA bearing viruses.
We interpret lack of reassortment within the PB1-PB2-HA complex across the Vic--Yam lineage boundary in recent times as evidence of a considerable degree of co-dependence within the Vic and Yam PB1-PB2-HA lineages, whereby recent reassortants bearing mixed-lineage PB1-PB2-HA complexes are inviable or unsampled by surveillance due to being outcompeted by viruses with pure-lineage PB1-PB2-HA complexes.

We have identified amino acid sites which exhibit a high degree of linkage disequilibrium between PB1, PB2 and HA segments.
Sites with high LD on PB1, PB2 and HA segments could be interpreted as either sites which have drifted apart due to lack of reassortment between the three segments or as sites which prevent the cooperation of the three proteins at the molecular level and thus prevent inter-lineage PB1-PB2-HA reassortants from having comparable fitness to viruses with `pure' lineage PB1-PB2-HA complexes.

@@

\subsection*{All segments reassort equally but some are more equal than others}
The 95\% HPD intervals of normalized approximate SPR distances between pairs of segments (Figure \ref{SPRdistances}) overlap, suggesting there is no evidence of differences in the numbers of reassortments (or reassortment rate given as number of SPR moves per tree, data not shown) between all segments.
There is reason to suspect that our SPR distance analysis simply lacks power.
We used an approximation of the true SPR distances and SPR distances themselves could only approximate (and underestimate) the actual numbers of reassortments.
However, exact SPR distances could be estimated for a limited number of segment pairs (PB1--PB2 and PB1--HA) and analyses indicate that after normalization exact and approximate SPR distances are not significantly different (supplementary figure of exact-approximate SPR distance differences and/or correlations of the two?).

Significant differences between pairs of segments can be found when considering `reassortment distance'.
Our estimates suggest that tips in PB1, PB2 and HA trees have similar TMRCAs (Figure \ref{deltaTMRCA}), suggesting that reassortments between the three rarely bring together lineages that have been evolving separately for prolonged periods of time.
We cannot determine whether this effect is specifically due to non-reassortment of Vic and Yam PB1-PB2-HA lineages or a generic property of these three segments.
In addition, we see evidence of this in MP and NA trees (comment - get references for MP-NA association in fluA, otherwise put down as personal communication or commonly held suspicion).

@@

It is perhaps easiest to explain co-dependence between PB1 and PB2 segments based on their functions as part of the influenza RNA-dependent RNA polymerase (RdRp) heterotrimer.
Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar trees.
In addition, PB1--PB2 reassortants are the rarest and least persistent amongst mixed-lineage PB1-PB2-HA strains and have not been isolated since 1996.

Explaining the co-dependence between PB1+2 and HA segments is more difficult.

@@

Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of antigenic diversity within the influenza B population (the Vic lineage NA segment having gone extinct in 2002), until the Yam lineage NA segment under Vic PB1-PB2-HA recovers sufficient antigenic diversity.
Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments', which show much more recent TMRCAs and less variability at the amino acid level \cite{chen2006,obenauer2006}.
It is also possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
This process, termed mutation-driven co-evolution \cite{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \cite{lee2008}.

Though it is possible that PB1 and PB2 segments are co-adapted and the HA segment is simply hitchhiking along stochastically, we find it unlikely given that B/Waikato/6/2005-like strains, the most recent isolates with a mixed-lineage PB1-PB2-HA complex, circulated for a limited amount of time within a limited geographic area, suggesting that this particular lineage was outcompeted by influenza B strains with `pure' PB1-PB2-HA complexes.

\subsection*{The future of influenza B viruses}
We see two potential paths of evolution for influenza B viruses.
More segments could be recruited into co-adapted segment complexes (PB1, PB2 and HA segments having started this first, akin to genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.

This may seem to be the case in Figure \ref{tmrcaOT} where the NP segment, exhibiting the fourth oldest TMRCA by 2011, looks like it might begin co-evolving with the PB1-PB2-HA complexes. 
However, the continuous measure of diversity between Vic and Yam lineages of PB1, PB2 and HA segments in Figure \ref{betweenDiversity} show that the NP segment is skewed towards very recent TMRCAs, suggesting that a large number of recent isolates are products of recent reassortment between the NP segment and PB1-PB2-HA complexes.
In fact, this is due to viruses possessing B/Brisbane/33/2008-like genome constellations, which comprise a large part of recently isolated influenza B virus strains (Figure \ref{railroadPlot}).
The other alternative is continued reassortment of all non-PB1-PB2-HA segments between Vic and Yam PB1-PB2-HA complexes, or even extinction of one or the other lineage of the PB1-PB2-HA complex.

Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether such two-lineage dynamics have not occurred in the history of influenza B viruses.
If the influenza B population is on its way to speciation we expect that over time inter-lineage reassortment frequency should decrease, as more and more segment lineages would be recruited into co-adapted segment complexes.

@@

Given the existence of B/Waikato/6/2005-like viruses and the rarity of PB1-PB2 reassortants we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1+2-HA reassortants.

@@

\section*{Conclusions}
In this study we have investigated the patterns of reassortment amongst two influenza B virus segment lineages and found preserved diversity and consistent co-reassortment of 3 segments: PB1, PB2 and HA.
Our findings reveal that Vic and Yam lineages of PB1, PB2 and HA segments co-associate amongst themselves with segments of their own lineage, thus resulting in an entirely Vic-lineage-derived PB1-PB2-HA segment complex and an entirely Yam-lineage-derived PB1-PB2-HA segment complex that co-circulate within the influenza B virus population, occasionally exchanging non-PB1-PB2-HA segments.
Though we detect strains with mixed-lineage PB1-PB2-HA complexes, we find that they have performed poorly and have circulated for short amounts of evolutionary time when compared to viruses with `pure' lineage PB1-PB2-HA complexes.

We argue that this is evidence for a selectively maintained relationship between PB1-PB2-HA complex of Vic and Yam lineages and not due to biased reassortment.
Given sufficient time we expect the two PB1-PB2-HA complexes to recruit the rest of the genomic segments into co-adapted and co-reassorting segment complexes eventually resulting in sympatric speciation of influenza B viruses.
